---
title: "SORA Inclusive Analytics Workshop"
output: html_notebook
---

Some R packages we will be using for the workshop. Feel free to add others during the session. 

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(dplyr)
library(RSocrata)
library(ggplot2)
library(inspectdf)
library(forcats)
```

The dataset we will be using is the State Expenditures data set, which you can load into your R session by passing the URL to `RSocrata::read.socrata()`. 

About the data: 
<https://data.oregon.gov/Revenue-Expense/Agency-Expenditures-Multi-Year-Report/y9g9-xsxs/about_data>

This report provides information on expenditures (i.e., cash transactions/payments) for 
the agencies that utilize the Statewide Financial Management Application (SFMA) issued 
for the fiscal year 2024 (July 1, 2023 - June 30, 2024). 

# Purpose

Before defining our purpose we should consider... 

## Who might be interested in an analysis of expenditures? 

* Public employees  
* Agencies/agency heads  
* Public/taxpayers  
* Legislators  
* Vendors/Suppliers  
* Auditors  
* Analysts


## What are the benefits?

* Process improvements (internal and external)  
* Viewing outliers (identifying too small or too large of expenditures) 
* Trends, e.g. seeing the general pattern - what are the highest/lowest expenditures, what are the most common expenses, etc.
* Compliance, to ensure that we are following our procurement rules
* Opportunities for consolidating, getting better contracts
* Better understanding of where taxes go

## What are the risks? 

* Legislators may see the analysis and think that costs are too high and use it as rationale to cut budgets  
* Patterns may be revealed that harm reputations (institutional, individual, etc)

Purpose: 
Make the expenditures data set more accessible to those interested in the analysis by providing

Expense trends  

* Overall expenses by year  
* Expense by year by agency  
* Expense by category by agency  
* Expense by category by vendor  

Deep dive into outliers  

* High & Low expenditures  
* Expenses out of compliance with procurement statues/rules/guidance

# Viewing the data  

*We do not build an analysis around the data we have; we find the data for the analysis that we need!*

This chunk shows two ways we can load the data, one relies on an API via RSocrata, whereas the second option relies on exporting the data manually from 
Data.Oregon.Gov.

```{r load-expenditure-data, echo=TRUE}
# Using RSocrata
state_expenditures <- read.socrata("https://data.oregon.gov/Revenue-Expense/Agency-Expenditures-Multi-Year-Report/y9g9-xsxs")

# Using csv file stored locally
#state_expenditures <- read.csv(here("data", "Agency_Expenditures_–_Multi-Year_Report_20250827.csv"), stringsAsFactors = FALSE)

glimpse(state_expenditures)
```
## Data quality check

We'll clean-up some of the types that should be treated as characters. We'll also rename variables for improved readability.

```{r echo=TRUE}
# basic clean-up
state_expenditures_clean <- 
  state_expenditures %>% 
  mutate(across(c(agency , expend_class, budget_class), ~as.character(.))) %>% 
  rename(
    "agency_code" = agency,
    "agency_name" = agency_1,
    "budget_class_code" = budget_class,
    "buget_class_name" = budget_class_1,
    "expend_class_code" = expend_class,
    "expend_class_name" = expend_class_1
  )

glimpse(state_expenditures_clean)
```
We can use `inspectdf` to view the unique counts, and most common values for each of the categorical variables. 

```{r echo=TRUE}
# unique counts by category
state_expenditures_clean %>% 
  inspect_cat()
```

We can see that there are discrepancies between the agency, budget, and expenditure class codes since the total counts for these are different. 
Let's take a look at those. 

```{r identify-duplicates, echo=TRUE}
# duplicates for agency codes
state_expenditures_clean %>% 
  distinct(agency_code, agency_name) %>% 
  count(agency_code, sort = TRUE) %>% 
  filter(n > 1)

# duplicates for budget class codes
state_expenditures_clean %>% 
  distinct(budget_class_code, buget_class_name) %>% 
  count(budget_class_code, sort = TRUE) %>% 
  filter(n > 1)

# duplicates for expenditure class codes
state_expenditures_clean %>% 
  distinct(expend_class_code, expend_class_name) %>% 
  count(expend_class_code, sort = TRUE) %>% 
  filter(n > 1)
```

What states are included for vendors?

```{r echo=TRUE}
# show all states included
state_expenditures_clean %>% 
  count(vendor_st, sort = TRUE)
```
Interestingly, this extends beyond the US which we recognize by the 'XX', and several Canadian provinces as well. 

At this point we may decide that the data quality issues require engagement with the data owners of this asset to proceed with our analysis. 

Or we may be comfortable with proceeding!

# Exploratory Analysis

Viewing total expenses by year

```{r echo=TRUE}
state_expenditures_clean %>% 
  group_by(fiscal_year) %>% 
  reframe(total_expenses_in_billions = sum(expense)/1e9)
```

```{r echo=TRUE}
state_expenditures_clean %>% 
  group_by(fiscal_year) %>% 
  reframe(total_expenses_in_billions = sum(expense)/1e9) %>% 
  ggplot(aes(x = as.factor(fiscal_year), y = total_expenses_in_billions)) + 
  geom_col() +
  scale_y_continuous(labels = scales::label_currency()) +
  theme_classic() + 
  labs(
    y = "Total Expenses ($ Billions)", 
    x = NULL,
    caption = "\nData source:\n \'Agency Expenditures - Multi-Year Report\'\n As of December 17, 2024 from Data.Oregon.Gov"
  )
```


How about viewing by agencies with the top 5 highest expenses in 2024? 

```{r echo=TRUE}
top_agencies <- 
  state_expenditures_clean %>% 
  filter(fiscal_year == 2024) %>% 
  group_by(agency_name) %>% 
  reframe(total_expenses = sum(expense)) %>% 
  arrange(-total_expenses) %>% 
  slice(1:5) %>% 
  pull(agency_name)

summary_by_top_agencies <- 
  state_expenditures_clean %>% 
  filter(agency_name %in% top_agencies) %>% 
  group_by(fiscal_year, agency_name) %>% 
  reframe(total_expenses_in_billions = sum(expense)/1e9) %>% 
  mutate(
    agency_name = fct_reorder(agency_name, total_expenses_in_billions, .desc = TRUE)
  )
  
summary_by_top_agencies  %>% 
  ggplot(aes(x = as.factor(fiscal_year), y = total_expenses_in_billions, group = agency_name)) + 
  geom_col() +
  facet_wrap(~ agency_name, ncol = 3, scales = "free_y", axes = "margins") + 
  scale_y_continuous(labels = scales::label_currency()) +
  theme_classic() + 
  labs(
    y = "Total Expenses ($ Billions)", 
    x = NULL,
    caption = "\nData source:\n \'Agency Expenditures - Multi-Year Report\'\n As of December 17, 2024 from Data.Oregon.Gov",
    title = "Agencies with the Top Highest Expenditures in 2024",
    subtitle = "Axes for total expenses varies by agency to show differences in scale."
  )
```

8.27.2025  
Amelia L. Vargas